Automated Lensing Learner - I: An automated Strong Lensing Identification Pipeline

### 2017-03-28 ms.tex
Added the minor changes from Tom's comments

%-------------------------------------------------%
\begin{abstract} 
Gravitational lensing offers a direct probe of the underlying mass
distribution of lensing systems, a window to the high redshift
universe, and a geometric probe of cosmological models.  The advent of
large scale surveys such as the Large Synoptic Sky Telescope and
Euclid has prompted a need for automatic and efficient identification
of strong lensing systems.  We present (1) {\em (ALL) Automated
  Lensing Learner}, a strong lensing identification pipeline that will
be publicly released as open source software, and (2) a publicly
available mock LSST dataset with strong galaxy-galaxy lenses.  In this
first application of the pipeline, we employ a fast feature extraction
method, Histogram of Oriented Gradients (HOG), to capture edge
patterns that are characteristic of strong gravitational arcs in
galaxy-galaxy lensing.  We use logistic regression to train a
supervised classifier model on the HOG of HST- and LSST-like images.
We use the area under the curve (AUC) of a Receiver Operating
Characteristic (ROC) curve to assess model performance; $AUC=1.0$ is an
ideal classifier, and $AUC=0.5$ is no better than randomly guessing.
Our best performing models on a training
set of 10,000 lens containing images and 10,000 non-lens containing
images exhibit an AUC of 0.975 for an HST-like sample.  However, for
one orbit of LSST, our model only reaches an AUC of 0.625.  For
10-year mock LSST observations, the AUC improves to 0.809.  Model
performance appears to continually improve with the size of the
training set.  Models trained on fewer images perform better in
absence of the lens galaxy, the removal of which can decrease the
false positive rate due to the blending of light between source images
and the lens galaxy.  However, with larger training data sets,
information from the lens galaxy actually improves model performance.
Our results demonstrate an efficient and effective method for automatically
identifying strong lenses that captures much of the complexity of the
arc finding problem. The linear classifier both runs on a
personal laptop and can easily scale to large data sets on a computing
cluster, all while using existing open source tools.
\end{abstract}
%-------------------------------------------------%

@@

% Gravitational lensing as a geometric test of cosmology.  Also
% probes underlying matter distribution in galaxy clusters, sensitive
% to things like recent accretion and environment.
Gravitational lensing occurs when intermediate fluctuations in the
matter density field deflect light from background sources
\citep[see][for a review]{kneibandnatarajan_11}.  Strong gravitational
lensing can manifest as visible giant arcs magnifying high redshift
galaxies \citep{lyndsandpetrosian_86,gladders_etal03}, multiply imaged
quasars \citep{walsh_etal79}, multiply imaged galaxies
\citep{sharon_etal05}, and arclets \citep{bezecourt_etal98}.  Lensing
signatures probe the underlying dark matter distribution of the lens
\citep{warrenanddye_03}, and high redshift galaxy formation
\citep{allam_etal07}. Strong lenses also provide a geometric test of
cosmology via comparison of predicted arc abundances with observed
abundances \citep{kochanek_96,chae_03,linder_04}, and via time-delay
between signals from multiply imaged quasars
\citep{xli_etal12,suyu_etal14,suyu_etal16}.

@@

% Image identification works: visual, ``robot'' identification -
% ArcFinder, RingFinder..., and citizen science (space warps).  Many
% of these rely on some explicit aspect of the image's morphology
% (give example), filtering, ... ....  Specifics of the methodology
% how the instrument affects the strong lensing signal, and the
% objects of interest.
Over the last decade, infrastructure for both large scale visual and
automated image classification emerged. By nature, the human eye is
one of the best discriminators for image classification.  Visual arc
identification has been effective through the use of citizen science
platforms.  {\em SpaceWarps} is an example of citizen science based
image classification of strong lensing systems in Canada France Hawaii
Telescope Science (CFHTLS) telescope observations
\citep{marshall_etal16,more_etal16}.  These platforms are quite
successful for a dataset like CFHTLS; here, 3,000 candidate images
were identified in eight months, resulting in 89 final candidates.
However, future datasets like Euclid \citep{oguriandmarshall_10} and
LSST expect to find tens of thousands galaxy-scale strong lenses
\citep{pawase_etal12}.  The volume of upcoming data challenges the
scalability of a pure citizen science approach.

@@

% Interpreting TP and FP rates
Finally, we must take the class balance, or relative number of lens to
non-lens systems, into account when assessing a metric.  The
Purity-Recall (purity-completeness) metric is sensitive to the ratio
of lens and non-lens systems in the data.  Both of our training and
test data sets have $50\%$ lensed and $50\%$ unlensed images, which is
not expected in observations.  The ROC curve metric is insensitive to
the ratio, but is sensitive to the sampling.  Given alternative lens
and non-lens sample splittings, our true positive and false positive
rates in the ROC curves would stay the same, making the ROC curve a
more standard metric in the literature.  On the other hand, the ROC
curves show a representative rate for lens and source distributions
that are evenly sampled.  We do {\it not} expect this sampling to be
representative of what we might expect from a an observational survey.
We leave these additional challenges to future work.

>>>>

%-------------------------------------------------%
\begin{abstract} 
Gravitational lensing offers a direct probe of the underlying mass
distribution of lensing systems, a window to the high redshift
universe, and a geometric probe of cosmological models.  The advent of
large scale surveys such as the Large Synoptic Sky Telescope and
Euclid has prompted a need for automatic and efficient identification
of strong lensing systems.  We present (1) {\em (ALL) Automated
  Lensing Learner}, a strong lensing identification pipeline that will
be publicly released as open source software, and (2) a publicly
available mock LSST dataset with strong galaxy-galaxy lenses.  In this
first application of the pipeline, we employ a fast feature extraction
method, Histogram of Oriented Gradients (HOG), to capture edge
patterns that are characteristic of strong gravitational arcs in
galaxy-galaxy lensing.  We use logistic regression to train a
supervised classifier model on the HOG of HST- and LSST-like images.
We use the area under the curve (AUC) of a Receiver Operating
Characteristic (ROC) curve to assess model performance; $AUC=1.0$ is an
ideal classifier, and $AUC=0.5$ is no better than randomly guessing.
Our best performing models on a training
set of 10,000 lens containing images and 10,000 non-lens containing
images exhibit an AUC of 0.975 for an HST-like sample.  However, for
one exposure of LSST, our model only reaches an AUC of 0.625.  For
10-year mock LSST observations, the AUC improves to 0.809.  Model
performance appears to continually improve with the size of the
training set.  Models trained on fewer images perform better in
absence of the lens galaxy, the removal of which can decrease the
false positive rate due to the blending of light between source images
and the lens galaxy.  However, with larger training data sets,
information from the lens galaxy actually improves model performance.
Our results demonstrate an efficient and effective method for automatically
identifying strong lenses that captures much of the complexity of the
arc finding problem. The linear classifier both runs on a
personal laptop and can easily scale to large data sets on a computing
cluster, all while using existing open source tools.
\end{abstract}
%-------------------------------------------------%

@@

% Gravitational lensing as a geometric test of cosmology.  Also
% probes underlying matter distribution in galaxy clusters, sensitive
% to things like recent accretion and environment.
Gravitational lensing occurs when intermediate fluctuations in the
matter density field deflect light from background sources
\citep[see][for a review]{kneibandnatarajan_11}.  Strong gravitational
lensing can manifest as visible giant arcs magnifying high redshift
galaxies \citep{lyndsandpetrosian_86,gladders_etal03}, multiply imaged
quasars \citep{walsh_etal79}, multiply imaged galaxies
\citep{sharon_etal05}, and arclets \citep{bezecourt_etal98}.  Lensing
signatures probe the underlying dark matter distribution of the lens
\citep{warrenanddye_03}, and high redshift galaxy formation
\citep{allam_etal07}. Strong lenses also provide a geometric test of
cosmology via comparison of predicted arc abundances with observed
abundances \citep{kochanek_96,chae_03,linder_04}, via time-delay
between signals from multiply imaged quasars
\citep{xli_etal12,suyu_etal14,suyu_etal16,bonvin_etal17}, and via
distance ratios in lenses with sources at multiple redshifts
\citep{jullo_etal10,collett_etal12,collettandauger_14}.

@@

% Image identification works: visual, ``robot'' identification -
% ArcFinder, RingFinder..., and citizen science (space warps).  Many
% of these rely on some explicit aspect of the image's morphology
% (give example), filtering, ... ....  Specifics of the methodology
% how the instrument affects the strong lensing signal, and the
% objects of interest.
Over the last decade, infrastructure for both large scale visual and
automated image classification emerged. By nature, the human eye is
one of the best discriminators for image classification.  Visual arc
identification has been effective through the use of citizen science
platforms.  {\em SpaceWarps} is an example of citizen science based
image classification of strong lensing systems in Canada France Hawaii
Telescope Science (CFHTLS) telescope observations
\citep{marshall_etal16,more_etal16}.  These platforms are quite
successful for a dataset like CFHTLS; here, 3,000 candidate images
were identified in eight months, resulting in 89 final candidates.
However, future datasets like Euclid \citep{oguriandmarshall_10} and
LSST expect to find hundreds of thousands galaxy-scale strong lenses
\citep{collett_15}.  The volume of upcoming data challenges the
scalability of a pure citizen science approach.

@@

% Interpreting TP and FP rates
Finally, we must take the class balance, or relative number of lens to
non-lens systems, into account when assessing a metric.  The
Purity-Recall (purity-completeness) metric is sensitive to the ratio
of lens and non-lens systems in the data.  Both of our training and
test data sets have $50\%$ lensed and $50\%$ unlensed images, which is
not expected in observations.

% Realistic ratios
For images selected for a massive elliptical, the number of non-lenses
will outnumber lenses by at least 1000 to 1.  Therefore, a sample that
is $50\%$ pure requires a classifier with a false positive rate of
0.001.  Looking at the green line in Figure~\ref{fig:PRcompilation}, a
classification threshold for a pure sample only encapsulates
$\sim10\%$ (in recall) of all of our lenses before contamination from
non-lenses leaks into the selection.  A maximally large and pure
sample from our method would require further filtering, e.g. by
citizen science.

% ROC curve metric
The ROC curve metric is insensitive to the ratio, but is sensitive to
the sampling.  Given alternative lens and non-lens sample splittings,
our true positive and false positive rates in the ROC curves would
stay the same, making the ROC curve a more standard metric in the
literature.  On the other hand, the ROC curves show a representative
rate for lens and source distributions that are evenly sampled.  We do
{\it not} expect this sampling to be representative of what we might
expect from a an observational survey.  We leave these additional
challenges to future work.
